Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-06-24 manuscript.qmd
finish risk figure, update some captions

model = shared.get_model(which="simple", load_samples=True)
samples = shared.get_samples(which="simple")

@@

![Comparison of predicted (histograms) vs observed (beta posteriors) prevalences. Shown are four scenarios, all including contralateral LNL II involvement: Early T-category and an ipsilateral N0 neck (green), early T-category and ipsilateral LNL II involvement (blue), as well as the same two scenarios but for advanced T-category (orange and red). The figure shows that the model is capable of describing the correlation between ipsi- and contralateral involvement. Although for the scenario of LNL II involvement in both sides, the prediction's split between early and advanced T-category is not large enough.](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}

In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement. It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side. However, there are some small discrepancies in the model's prediction: When considering the scenario of LNL II involvement in the ipsi- *and* contralateral side, it cannot quite capture the split between early and late T-category. The model slightly overestimates the prevalence of this scenario for early T-category patients and underestimates it for advanced T-category.

@@

![Histograms over the predicted risk of occult contralateral LNL II invovlement, shown for some combinations of T-category, tumor lateralization, and ipsilateral clinical involvement. The contralateral side was always assumed to be clinically negative.](figures/fig-model-risks.png){#fig-model-risks width=50%}

@Fig-model-risks shows this predicted risk of occult disease in contralateral LNL II for four interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].

The variable impacting the prediction for contralateral involvement in our model is the tumor's lateralization. For example, a patient with a clearly lateralized early T-category tumor and a clinically N0 neck is assigned a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II. Under the same scenario, but *with* mid-sagittal extension, the risk jumps to almost 7%. <!-- TODO: make this dynamic -->

@@

>>>>

model = shared.get_model(which="full", load_samples=True)
samples = shared.get_samples(which="full")

@@

![Comparison of the computed and observed prevalences for scenarios that illustrate the model's capability of accounting for the correlation between ipsi- and contralateral involvement. We show two scenarios: One where the ipsilateral neck shows no involvement (at least LNLs I to V are healthy, LNL VII was ignored because data on it is missing for some patients) in blue and one where at least LNL III was involved in orange. These two scenarios are plotted for all combinations of T-category (early in top row, advanced in bottom row) and tumor lateralization (lateralized in left column, extending over mid-sagittal line in right column).](figures/fig-model-prevalences-with-ipsi.png){#fig-model-prevalences-with-ipsi}

In @fig-model-prevalences-with-ipsi we display the model's ability to capture the correlation between ipsi- and contralateral involvement. It shows that the prevalence of metastases in the two sides of the neck is correlated via the time of diagnosis, despite the model not having any direct connections between the two side. However, there are some discrepancies in the model's prediction: It cannot quite capture the correlation of ipsi- and contralateral involvement that is seen in the data. This is notable for example in the bottom right subplot where the prevalence is overestimated for the scenario of an (almost) healthy ipsilateral neck and underestimated when at least LNL III shows metastases. One reason for this may be the width of the time-prior: If it was wider, the posterior over the diagnosis time given the ipsilateral involvement could more flexibly shift to reflect a more severe contralateral involvement.

<!-- NOTE: This last sentence is purely speculative! We need to discuss this. -->

@@

![Histograms over the predicted risk of occult contralateral LNL II (left), III (middle), and IV (right) invovlement, shown for some combinations of T-category, tumor lateralization, and ipsilateral clinical involvement. Unless otherwise mentioned in the legend, the entire contralateral side was assumed to be clinically node negative.](figures/fig-model-risks.png){#fig-model-risks}

@Fig-model-risks shows this predicted risk of occult disease in the contralateral LNLs II, III, and IV for interesting combinations of these three risk factors. This risk is computed *given* a CT diagnosis assessing the per-LNL clinical involvement for which we assumed a sensitivity of 81% and a specificity of 76% [@de_bondt_detection_2007].

### Contralateral LNL II

The variable impacting the prediction for contralateral level II involvement in our model is the tumor's lateralization. For example, a patient with a clearly lateralized early T-category tumor and a clinically N0 neck is assigned a 1-2% <!-- TODO: make this dynamic --> risk for occult disease in contralateral LNL II. Under the same scenario, but *with* mid-sagittal extension, the risk jumps to almost 7%. <!-- TODO: make this dynamic -->

@@

### Contralalteral LNL III

As shown in the center subplot of @fig-model-risks, the risk for occult disease in the contralateral LNL III only ever crosses a 5% threshold if the upstream level II is involved too, resulting in a risk of 5% to 6%<!-- TODO: make dynamic -->. As long as the contralateral neck appears clinically node negative, however, even an advanced T-category and midline extension do not yield a risk higher than 2%.

### Contralateral LNL IV

Our model's prediction for the contralateral LNL IV show that even in the most extreme case, with an advanced T-category tumor extending over the mid-sagittal line as well as metastases in every ipsilateral LNL and every contralateral upstream level, the risk does not exceed 5%
